\documentclass[../../what-is-computer.tex]{subfiles}
\begin{document}
    Let's once again refer to Merriam-Webster dictionary for a \emph{computer} definition\footnote{\href{https://www.merriam-webster.com/dictionary/computer}{https://www.merriam-webster.com/dictionary/computer}}:

    \begin{quote}
        computer is a programmable usually electronic device that can store, retrieve, and process data.
    \end{quote}

    So, a computer directly tied to all sorts of \emph{data maniplation}. But what is data, and how can it be manipulated? Data is pretty much any factual information.
    Weather outside a window? Data. T-Shirt you're wearing? Data. Dusty books in my shelves? Data. But there are a certain layers present. The fact that I'm wearing a T-Shirt 
    is data (even if I'm not --- also data). What color it is is also, most certainly data. Is there any text or image present?
    Both the existence and \emph{information} in it --- also data. Let's also not forget about colors, sizes, fonts. We are living and have always lived in an enormous ocean of data. Nowadays, with our technology even more so. \par

    But do we have use for this data? Well, the answer is --- it depends. To assess data without well-defined goal will almost always result in you drowning in said data without much progress,
    since world offers us practically indefinite source of it. We must put our data into some perspective, some context. Once we put our data in some context and it becomes \emph{useful} for us, \emph{it becomes information}.

    \begin{quote}
        To somehow navigate in this world, we must put our data in context. Data with given context becomes somewhat useful. Such data called \emph{information}.
    \end{quote}

    Information is much more well defined than data. \emph{We can measure information, actually}. We even have a science discipline, called \emph{Information theory},
    that researches all about information, from mathematical and engineering standpoint. Not only that, but we have an entire industry, called \emph{Information Technologies}
    built entirely on a foundation provided by Information theory and adjacent disciplines.\par

    But until we dwell into technical stuff, we must also remember, that \emph{we used to operate with information}. Our brain is a natural computer, disecting data
    into information that we use in our everyday life. How come I am so sure to call that information and not just raw data? Well, that greatly depends on a 
    scale, but
    \emph{our brain have very defined goals}. One of the main said goals being to \emph{keep us alive}. Therefore there is a context, and brain will always tend to 
    categorize things (organize data) based on this goal\footnote{If your goal at the moment being something more specific, than just stay alive, many of the information brain gives you still raw data},
    making it somewhat useful, therefore making it an \emph{information}. Even the simple fact, that we don't notice our noses, although our eyes do see it constantly, 
    tells us how natural brain is in working with information. \par

    So, being a natural organic computer, we must first understand what we do with information, to have insight on what computers do with information. All processing 
    of information, our minds in work is mostly an internal process. \emph{Sooner or later there is a point, when we must exchange said information}. So, how do we
    exhange it? We have an enormous number of ways, actually. We can say something to another person, write it down, pass via somebody a note, we can just hint at something.
    We can send message in a messenger, send an email, send a radio-signal, we can knock a Morse code. \emph{We are practically limitless with one major nuance --- the other
    side must be able to understand us}. There is no point in sending email to a person, who can't use it\footnote{In general. Sometimes we are interested only
    in sending information, not concerned by an actual delivery. Some legal procedures can be of an example}. \par

    We now can conclude one fundamental distinction: information itself is mostly independent from it's carrier. To demonstrate it more clearly, let us consider
    following example: I want to send information to my friend at the table, with the main message being \emph{`pass me salt'}. There is definetely a context:
    we are at the table, eating food, there are at least two parties involved (me and my friend), and I expect salt to exist somewhere at the table. I can pass this 
    information with a various number of ways, provided my friend understands me. Just to mention a few:

    \begin{itemize}
        \item Say to him `pass me salt'
        \item Ask him to pass me salt in other language, he is familiar with
        \item Write a note to him `pass me salt'
        \item Write a note in foreign language, he is familiar with
        \item Get his attention and non-verbally point at salt
        \item Write him a message in messenger, expecting phone to be near them
        \item Get his attention and use ASL or alternative, provided he is familiar with it
        \item Exclaim obviously `Oh! This food will be so much better with salt! I wish somebody passed it to me now', provided he understood our hint
        \item Rhytmically knock with Morse code, provided he understands it
    \end{itemize}

    In all aforementioned examples we can clearly see, that the gist of our `message' stayed the same. \emph{We did pass a more or less the same information} in 
    each and every case. Despite the medium being completely different, if our friend can undestand us, nothing really changed for him or us. In such cases
    the `main message' containing an actual useful information we are willing to exchange usually colloquially called a \emph{`payload'}. However,
    despite our `payload' being virtually the same, we did pass some additional information (or data --- depending on context) along the way, didn't we?

    Let's use an above list one more time, but will provide additional few details, just for example:

    \begin{itemize}
        \item Say to him `pass me salt'
        \begin{itemize}
            \item In what voice tone?
            \item How loud did we ask?
            \item What face expressions followed along our request?
            \item Was there any gesticulation involved? How intense?
            \item In what speed we asked?
        \end{itemize}

        \item Ask him to pass me salt in other language, he is familiar with
        \begin{itemize}
            \item What language?
            \item Was there any context in using this language?
            \item In what voice tone?
            \item How loud did we ask?
            \item What face expressions followed along our request?
            \item Was there any gesticulation involved? How intense?
            \item In what speed we asked?
        \end{itemize}

        \item Write a note to him `pass me salt'
        \begin{itemize}
            \item What font did we use?
            \item Is it hand-written?
            \item Font color?
            \item Font size?
            \item Paper type?
            \item Paper size?
            \item Was paper plain white, or was it with pictures?
        \end{itemize}

        \item Write a note in foreign language, he is familiar with
        \begin{itemize}
            \item What language?
            \item Was there any context in using this language?
            \item What font did we use?
            \item Is it hand-written?
            \item Font color?
            \item Font size?
            \item Paper type?
            \item Paper size?
            \item Was paper plain white, or was it with pictures?
        \end{itemize}

        \item Get his attention and non-verbally point at salt
        \begin{itemize}
            \item Were we mumbling at the same time?
            \item How we got his attention? Tapped his shoulder? How strongly?
            \item How did we point? With a finger, palm, node?
            \item What face expressions have we used?
            \item How fast did we do it? 
        \end{itemize}

        \item Write him a message in messenger, expecting phone to be near them
        \begin{itemize}
            \item Have he seen us typing a message?
            \item How fast he reacted?
            \item Was there any notification
            \item Did we send an emoji?
            \item Did we send some attachment?
            \item We sent one message or several?
            \item Did he read it?
        \end{itemize}

        \item Get his attention and use ASL or alternative, provided he is familiar with it
        \begin{itemize}
            \item How exactly did we phrase our request? By letters or by gests?
            \item How fast did we transmit?
            \item We followed along with our lips?
        \end{itemize}

        \item Exclaim obviously `Oh! This food will be so much better with salt! I wish somebody passed it to me now', provided he understood our hint
        \begin{itemize}
            \item What intonation did we use?
            \item How loud did we say it?
            \item Where to did we look?
            \item Do we have any specific accents?
            \item Was there an emphasis on some words?
        \end{itemize}

        \item Rhytmically knock with Morse code, provided he understands it
        \begin{itemize}
            \item What period of time we used as an interval?
            \item Did we repeat our message? How many times?
            \item On what surface did we transmit?
            \item Did we use our knuckle? Spoon? Knife?
        \end{itemize}
    \end{itemize}

    So, we can conclude, that despite our \emph{payload} being practically the same, we did pass additional information along with it. To put it into perspective,
    It's somewhat similar, as if we were asked to describe an envelope, it's size, stamps on it and additional notes, ignoring the payload, being a letter inside
    said envelope. Such auxiliary information, which is more often than not isn't of our interest, however \emph{can be useful in certain scenarios}. Such data
    usually describe optional information aboout the \emph{payload} itself, or details of how it was delivered. Such data usually called \emph{metadata}

    \begin{quote}
        The information itself, that we wish to store or exchange colloquially called a \emph{payload}. Some additional details, that might be useful, regarding
        that information, but not tied to it directly usually called \emph{metadata}
    \end{quote}

    Let's say, I am writing a document in Microsoft Word. The \emph{payload} here being anything, I typed directly in this document. However, once I saved it, not
    only the document itself was saved, but also a bunch of additional \emph{metadata}. It can include date of the document creation, 
    author\footnote{Usually currently active user on OS level} of the document, last save date, last print date, etc.\emph{The same logic is applicable for virtually
    any file you've ever created}\footnote{Sometimes, ability to control metadata becomes crucial to save sensitive and personal information. 
    Some professions can put you in physical danger, if you're not cautious enough}.\par

    \newpage
    \section{You can't manage what you can't measure}
    Once we have gotten ourselves these neat definitions about data and information, we will not benefit from them until we resolve one fundamental issue:
    \emph{How do we measure information?} How can we possibly find an adequate solution to count something so abstract in nature. \par 

    Well, first of all this is where this fine distinction between \emph{data} and \emph{information} comes in play. You see, \emph{defining context} to transform
    our data into information provided us with one \emph{fundamental advantage}. To demonstrate this, let's consider an example: I have a drawer, where a bunch
    of my T-Shirts is stored. Let's say, there is no order whatsoever and once I open the drawer, any of my shirts can be on top\footnote{I would call
    that scenario uncanningly realistic}. Let's say that I for a fact know that there is no more than 12 T-Shirts of different colors totally in my drawer.
    I open a drawer and see a green T-Shirt on top. \emph{How much information did I receive}? \par

    To answer this question, let's investigate how any data is measured, for starters. Well, universally in information theory there is a number of data 
    measurement nomenclatures. The most ubiquitous and universal one being \emph{bits}. It got it's name from a `binary digit'. The binary part represents a
    tiny set of values it can be equal to, consisting only of \{0, 1\}.

    \begin{quote}
        Bit is a unit of information. It can be equal to either `1' or `0'.
    \end{quote}

    Bits are the most basic and fundamental part of any computer-related information operation. \emph{Everything} on your computer is stored in bits. Yes, 
    \emph{everything}. Everything you type in your Word documents, PowerPoint presentations, every audio track you've ever played, every site you've ever 
    visited, every file you've created, \emph{everything}. This simple fact might leave us with a bit of confusion: \emph{How on Earth can we measure apparently
    everything with such basic unit, only accepting `0' and `1' as a value?}. The answer is -- well, pretty easily. Once we have something, that cannot possibly
    be presented as value in a set of \{0, 1\}, \emph{we just increase the number of bits to store that additional information}. \par

    Consider 1 bit. Total number of values it can represent is limited to 2, by definition. It's either `1' or `0'. Well, if we increase the total numbers
    of bits and consider we have 2 bits. Now we have 2 possible `cells', each of which can `hold' 2 possible variants. Those values being: `00', `01', `10', `11'.
    Now, if something is still bigger than one of four possible variants, we can add bits until it will finally be enough. With a little bit of discrete math
    we can say for sure, that total possible variants a sequence of bits can hold equals to $2^x$, where 2 is a number of possible options a bit can `hold' (`0', `1'), 
    and $x$ is the number of bits we are ready to \emph{allocate}.

    \begin{table}[h]
        \centering
        \begin{longtable}{|c|c|c|}
            \hline
            Number of bits & Maximum options & Options \\\hline
            1 & 2 & \{0, 1\} \\\hline
            2 & 4 & \{00, 01, 10, 11\} \\\hline
            3 & 8 & \{000, 001, 010, 011, 100, 101, 110, 111\} \\\hline
            \ldots & \ldots & \ldots \\\hline
            $x$ & $2^x$ & \{ $x$ times 0, \ldots, $x$ times 1\} \\\hline
            \caption{Total possible permutations of bits}

        \end{longtable}
    \end{table}

    To count something in bits is pretty similar as to how we would count something normally, but with one nuance. In a modern world, humans mostly calculating 
    everything in a $base_{10}$. This simply means, that we have \emph{10 digits} that construct \emph{all of our numbers}\footnote{There are examples of cultures, that used different number as their base. However $base_{10}$ is the ubiquitous one nowadays}. 
    Those digits being: \{0, 1, 2, 3, 4, 5, 6, 7, 8, 9\}. So, our \emph{digits} can represent one of at-most 10 options. But how can we describe something more
    than 9? Well\ldots{} just start over and add additional digit! This is the same principle we follow, when we are counting in a $base_2$, working with bits.
    We can use pretty much any base we want, but along with $base_2$, $base_8$ and $base_{16}$ can be seen used widely\footnote{Writing everything in $base_2$
    can be slightly inconvenient, when exchanging information with other programmers. Those bases gives us ability to `shorten' binary code. Since bases
    are $2^3$ and $2^4$ respectively, we can `group' together bits in a group of 3 (in the case of $base_8$) or 4 (in the case of $base_{16}$), making it 
    easier for a human reading. Computer still operates with them in `raw' $base_2$ format}. So we can pretty much convert number in any base to any other 
    base, always keeping consistency:

    \begin{table}[H]
        \centering
        \begin{longtable}{|c|c|c|c|}
            \hline
            $base_{10}$ & $base_2$ & $base_8$ & $base_{16}$ \\\hline
            $0_{10}$ & $0_2$ & $0_8$ & $0_{16}$ \\\hline
            $1_{10}$ & $1_2$ & $1_8$ & $1_{16}$ \\\hline
            $2_{10}$ & $10_2$ & $2_8$ & $2_{16}$ \\\hline
            $3_{10}$ & $11_2$ & $3_8$ & $3_{16}$ \\\hline
            $4_{10}$ & $100_2$ & $4_8$ & $4_{16}$ \\\hline
            $5_{10}$ & $101_2$ & $5_8$ & $5_{16}$ \\\hline
            $6_{10}$ & $110_2$ & $6_8$ & $6_{16}$ \\\hline
            $7_{10}$ & $111_2$ & $7_8$ & $7_{16}$ \\\hline
            $8_{10}$ & $1000_2$ & $10_8$ & $8_{16}$ \\\hline
            $9_{10}$ & $1001_2$ & $11_8$ & $9_{16}$ \\\hline
            $10_{10}$ & $1010_2$ & $12_8$ & $A_{16}$ \\\hline
            $11_{10}$ & $1011_2$ & $13_8$ & $B_{16}$ \\\hline 
            $12_{10}$ & $1100_2$ & $14_8$ & $C_{16}$ \\\hline
            $13_{10}$ & $1101_2$ & $15_8$ & $D_{16}$ \\\hline
            $14_{10}$ & $1110_2$ & $16_8$ & $E_{16}$ \\\hline
            $15_{10}$ & $1111_2$ & $17_8$ & $F_{16}$ \\\hline
            $16_{10}$ & $10000_2$ & $20_8$ & $10_{16}$ \\\hline 
            \ldots & \ldots & \ldots & \ldots \\\hline
            \caption{Numbers in different bases}

        \end{longtable}
    \end{table}


    This concept was wonderfully explained in Charles Petzold's book \emph{Code: The Hidden Language of Computer Hardware and Software}. He explains different
    number bases with a following concept: we, humans, use $base_{10}$ mostly because it's very conveniently translates to \emph{number of our fingers and toes}.
    So, a $base_2$ would be only natural, if math was invented by dolphins! Since they have only two flippers from each side it would be super convenient for
    them to calculate. \par

    \begin{quote}
        To convert $base_{10}$ number to $base_{2}$ number, you can consider following method:
        \begin{enumerate}
            \item Divide your number by 2. You need to store both the \emph{quotient} end \emph{remainder}.
            \item Take your quotient as a number and repeat steps 1--2, until you got zero as a quotient
            \item Write down all your remainder in \emph{reverse order}. This is your binary number.
        \end{enumerate}

        e.g.\\
        We want to transform $35_{10}$ into binary form:
        \begin{table}[h]
            \centering                
            \begin{longtable}{|c|c|c|c|}
                \hline
                Step \# & Number & Quotient & Remainder \\\hline
                1 & 35 & 17 & 1 \\\hline
                2 & 17 & 8 & 1 \\\hline
                3 & 8 & 4 & 0 \\\hline
                4 & 4 & 2 & 0 \\\hline
                5 & 2 & 1 & 0 \\\hline
                6 & 1 & 0 & 1 \\\hline
                \caption{Step-by-step operations to convert $35_{10}$ to $base_2$}

            \end{longtable}
        \end{table}

        Now we got a list of \emph{remainders}: \{1, 1, 0, 0, 0, 1\}. Once we put them in \emph{reverse} order, we got our binary form: $35_{10} = 100011_2$
    \end{quote}


    One might wonder: But why choose $base_2$ number system for computers in the first place? Well, the answer is, like many things, a result of overcoming
    real-world limitations. Any computer utilises some hardware, tangible, physical metal parts. Since computers nowadays are mostly electronic devices 
    it made sense to create system detecting presence or absence of some electric signal. Presence of said signal would be connoted as `1` and absence as `0'. \par

    Now, since we dwelled enough into subject, it's time to answer to the initial question of the subsection: \emph{How much data did I receive, seeing a green 
    T-Shirt on the top, once I opened my drawer}. We can even add some precision to this question, asking not \emph{How much data \ldots}, but \emph{How many bits of information \ldots}.
    Well, the fundamental \emph{advantage} of data being in context, that I claimed in the beginning of the section is that now we know \emph{total number of options}.
    Since we know, that I only have 12 T-Shirts (and we don't really care about any other clothing in the drawer), there 12 possible options this could've ended. To 
    conclude the total number of bits this information will occupy, I can use something called \emph{Hartley function} to find out. It can be 
    represented\footnote{This form is somewhat simplified for the sake of clarity. Full Hartley Formula can be denoted as: $H_0(A) := \log_b|A|$}
    as: $2^x \geq N$ with $N$ --- representing our total outcomes number, 2 --- representing possible options for one bit and 
    \emph{$x$ --- number of bits we need to store it}.
    Mind `\geq' here --- since bit is indivisible measurement, we cannot just take 1.5 bits, for example.
    \emph{All cases ending up with a non-integer number need to be rounded up}. \par

    Since we know our total number of possible outcomes --- 12, we can calculate total numbers of bits we need to store that information, being the nearest power of 2
    exceeding possible outcomes number.

    \begin{quote}
        $2^4 \geq 12$ $\implies$ we need \emph{4 bits} to store information about \emph{any} T-Shirt being on top.
    \end{quote}

    Please, mind an important detail: we need 4 bits \emph{regardless} of which exactly T-Shirt will end up on top. To represent which T-Shirt it was exactly
    in each case of me, opening my drawer (provided, that total number of T-Shirts is constant), we have to come up with a \emph{code}.

    \begin{quote}
        Code --- a system of signals or symbols for communication\footnote{\href{https://www.merriam-webster.com/dictionary/code}{https://www.merriam-webster.com/dictionary/code}}.
    \end{quote}

    We must assign to every one of my T-Shirts a special \emph{system of symbols}. Since information is measured in bits, it's only natural for our `symbols' to 
    be simple bits sequences. Number of bits in these sequences will be equal to 4, since $2^4 \geq 12$\\
    Let's come up with a system of codes for T-Shirts:

    \begin{table}[H]
        \centering
        \begin{longtable}{cc}
            \toprule
            T-Shirt & Code \\
            \toprule
            Red & 0000 \\
            Brown & 0001 \\
            Green & 0010 \\
            Yellow & 0011 \\
            Pink & 0100 \\
            Gray & 0101 \\
            Purple & 0110 \\
            Blue & 0111 \\
            Black & 1000 \\
            White & 1001 \\
            Orange & 1010 \\
            Beige & 1011 \\
            \bottomrule
            \caption{Our \emph{encoding} for T-Shirts}
        \end{longtable}
    \end{table}
    So, upon opening my drawer and seeing my green T-Shirt on top I received 4 bits of information. If my drawer (or the T-Shirt itself) could send me binary
    information, they would send me `0010' representing it's color in this particular context.\par

    \begin{quote}
        Thus, we created an \emph{encoding} for our T-Shirts.
    \end{quote}
\end{document}