\documentclass[../../../programming-introduction.tex]{subfiles}
\begin{document}
    Basically, \emph{everything} you do on a computer is running some program. Everything, no matter what you do is a result of executing and working with
    some program. Your OS\footnote{Operation System --- Windows, Linux-based, MacOS, Android, iOS etc.} \emph{is a program}. Your internet browser is a program. 
    Your media player, file explorer, video game, media editing software, office software \emph{are programs}. 
    \emph{Everything that allows you to interact\footnote{Interaction with a computer here and on will not take physical interaction with a bare metal in account} 
    with computer (and not just bare metal) is a program.} \par

    So, essentially, \emph{what is a program?}. According to Merriam-Webster\footnote{Here and on Merriam-Webster dictionary is referred to as a `general-scope' dictionary to avoid technical details redundant for this essay} 
    definition\footnote{\href{https://www.merriam-webster.com/dictionary/program}{https://www.merriam-webster.com/dictionary/program}} 
    (applicable for us), we can use following as a definition: \par

    \begin{quote}
        program is a sequence of coded instructions that can be inserted into a mechanism (such as a computer)
    \end{quote}

    The gist of it being --- \emph{sequense of instructions}. So every interaction we can possibly have with a computer is somehow just a set of instructions. But \emph{how computers
    do understand out instructions}? If I just shout into the microphone some command, computer will not just do as I say\footnote{Provided, there is no running program, responsible for such behavior}.
    The same effect will have some instruction that I carefully write them in some text document, using Microsoft Word, for example. \par
    How do I make computer to understand what I want from it? To understand this, \emph{we must first understand what is a computer}. \par
\end{document}